---
title: 'Lecture 4: Stan Demo + Multivariate Normal intro'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(rstan)
#setwd("/Volumes/GoogleDrive/My Drive/Teaching/STAT534/Lectures/Lecture 4. Stan and GP")
```

#### Stan Overview

Recall the weather data from Bridger Bowl in January of 2021.

```{r}
temp <- c(26, 45, 44, 36, 22, 25, 31, 31, 37, 34, 35, 37, 32, 31)
```

We are going to fit a few this model in stan but we will vary the prior distribution to assess how the posterior distribution and results change.

##### 1. Stan Code

First re-write the stan code to allow mu and sigma to be specified by the user and passed to stan.

```{stan, output.var="just_text", eval = F, echo = T}
data {
  int<lower=0> N;
  vector[N] y;
  real mu0;
  real<lower=0> sigma0;
}

parameters {
  real mu;
  real<lower=0> sigma;
}

model {
  y ~ normal(mu, sigma);
  mu ~ normal(mu0, sigma0);
}

```

##### 2. Compute Posterior with Different Priors

Use the model to estimate the mean high temperature, but using the different prior structures below:

1. mu ~ N(20,10^2)

```{r, results = F}
prior1 <- stan("bridger.stan", data=list(N = length(temp), y=temp, mu0 = 20, sigma0 = 10))
```

```{r}
print(prior1, pars = 'mu')
```

2. mu ~ N(20,1000^2)

```{r, include = F}
prior2 <- stan("bridger.stan", data=list(N = length(temp), y=temp, mu0 = 20, sigma0 = 1000))
```

```{r}
print(prior2, pars = 'mu')
```


3. mu ~ N(20,.1^2)

```{r, include = F}
prior3 <- stan("bridger.stan", data=list(N = length(temp), y=temp, mu0 = 20, sigma0 = .1))
```

```{r}
print(prior3, pars = 'mu')
```

4. Explore another prior distribution. If you want to challenge yourself

##### 3. Visualize Posterior and Prior

1. mu ~ N(20,10^2)

```{r}
tibble(sims = c(extract(prior1, pars = 'mu')$mu,rnorm(4000, 20, 10)), 
       Distribution = rep(c('posterior','prior'), each = 4000)) %>% 
  ggplot(aes(x = sims, color = Distribution)) + 
  geom_density() + theme_bw() + 
  xlab('Temperature (F)') + ylab('') + 
  ggtitle("Prior and posterior belief for winter temperature in Bozeman") +
  labs(caption = 'Setting 1. mu ~ N(20,10^2)') +
  geom_vline(xintercept = mean(temp)) +
  annotate('text',label = expression(bar(x)), x=41, y=.1) +
  annotate('segment', arrow = arrow(length = unit(.1, 'inches')), x = 40, y = .1, yend = .1, xend = mean(temp))
```


2. mu ~ N(20,1000^2)

```{r}
tibble(sims = c(extract(prior2, pars = 'mu')$mu,rnorm(4000, 20, 10)), 
       Distribution = rep(c('posterior','prior'), each = 4000)) %>% 
  ggplot(aes(x = sims, color = Distribution)) + 
  geom_density() + theme_bw() + 
  xlab('Temperature (F)') + ylab('') + 
  ggtitle("Prior and posterior belief for winter temperature in Bozeman") +
labs(caption = 'Setting 2. mu ~ N(20,10^2)') +
  geom_vline(xintercept = mean(temp)) +
  annotate('text',label = expression(bar(x)), x=41, y=.1) +
  annotate('segment', arrow = arrow(length = unit(.1, 'inches')), x = 40, y = .1, yend = .1, xend = mean(temp))  
```


3. mu ~ N(20,.1^2)

```{r}
tibble(sims = c(extract(prior3, pars = 'mu')$mu,rnorm(4000, 20, .10)), 
       Distribution = rep(c('posterior','prior'), each = 4000)) %>% 
  ggplot(aes(x = sims, color = Distribution)) + 
  geom_density() + theme_bw() + 
  xlab('Temperature (F)') + ylab('') + 
  ggtitle("Prior and posterior belief for winter temperature in Bozeman") +
labs(caption = 'Setting 3. mu ~ N(20,.1^2)') +
  geom_vline(xintercept = mean(temp)) +
  annotate('text',label = expression(bar(x)), x=41, y=3) +
  annotate('segment', arrow = arrow(length = unit(.1, 'inches')), x = 40, y = 3, yend = 3, xend = mean(temp))  
```


#### Multivariate Normal Distribution

Next we will segue from standard linear models to analyzing correlated data.

First we will start with the a bivariate normal distribution:
y ~ N(theta,sigma), where theta is a mean vector and sigma = sigmasq * I is a covariance matrix.

To provide a motivating context, not consider jointly estimating the temperature at Bridger Bowl *and* Big Sky Resort.


##### 1. Simulate independent bivariate normal

Simulate a set of temperature values from each location, where the temperature values are independent (sigma = sigmasq * I)

```{r}
library(mnormt)
n <- 100
theta <- c(15,25)
sigma <- diag(2) * 100
fake_temperatures <- rmnorm(n, theta , sigma)
```

Then create a few graphs to show marginal distribution of temperature as well as how the temperatures evolve in time.

```{r}
library(reshape2)
melt(fake_temperatures, value.name = 'temp') %>% 
  rename(location = Var2) %>%
  mutate(location = factor(location)) %>% 
  ggplot(aes(x =temp, fill = location)) + 
  geom_histogram() +
  facet_wrap(.~location) + theme_bw() 

melt(fake_temperatures, value.name = 'temp') %>% 
  rename(location = Var2, day = Var1) %>%
  mutate(location = factor(location)) %>%
  ggplot(aes(y =temp, x = day, color = location )) + 
  geom_line() + theme_bw() + xlim(0,30) + 
  ggtitle('First 30 observations of independent response')

```

##### 2. Simulate correlated bivariate normal

Simulate a set of temperature values from each location, where the temperature values are not independent (sigma = sigmasq * H), where H is a correlation matrix. (Note there are some constraints we will discuss later)

```{r}
sigma <- matrix(c(1, .9, .9, 1), nrow = 2, ncol = 2) * 100
fake_temperatures_corr <- rmnorm(n, theta , sigma)
```

Then create a few graphs to show marginal distribution of temperature as well as how the temperatures evolve in time.

```{r}
melt(fake_temperatures_corr, value.name = 'temp') %>% 
  rename(location = Var2) %>%
  mutate(location = factor(location)) %>% 
  ggplot(aes(x =temp, fill = location)) + geom_histogram() +
  facet_wrap(.~location) + theme_bw()

melt(fake_temperatures_corr, value.name = 'temp') %>% 
  rename(location = Var2, day = Var1) %>%
  mutate(location = factor(location)) %>%
  ggplot(aes(y =temp, x = day,color = location )) + 
  geom_line() + theme_bw() + xlim(0,30) +
  ggtitle('First 30 observations of correlated response')

```

##### 3. Write STAN code for bivariate normal

Write stan code that will allow you to estimate theta and sigma (including H)

```{stan output.var='multi_norm', eval = F}
data {
  int<lower=0> p;
  int<lower=0> N;
  matrix[N,p] y;
}

parameters {
  vector[p] theta;
  corr_matrix[p] H;
  real<lower = 0> sigma;
}

model {
  for(i in 1:N){
    y[i,:] ~ multi_normal(theta, sigma*H);
  }
}

```

##### 4. Use STAN to estimate bivariate normal parameters

Use your stan code to estimate theta and sigma (including H and sigmasq)

```{r, results = F}
indep_mvn <- stan("multi_norm.stan", 
                  data=list(N = nrow(fake_temperatures), 
                            p = ncol(fake_temperatures), 
                            y=fake_temperatures))
```

```{r}
print(indep_mvn)
```

```{r, results = F}
corr_mvn <- stan("multi_norm.stan", 
                 data=list(N = nrow(fake_temperatures_corr),
                           p = ncol(fake_temperatures_corr), 
                           y=fake_temperatures_corr))
```

```{r}
print(corr_mvn)
```

##### 5. Final Thoughts About Correlation

In many statistical models there is an assumption about independence. When independence is violated, uncertainty is under estimated and in incorrect inferences can be made.

While lack of independence often has a negative connotation, in spatial statistics we can actually exploit correlation. For instance, by knowing the temperature at the weather station at Bozeman High School or Bridger Bowl, we can estimate temperature at other locations.
